---
title: "Exploratory Analysis"
output: html_document
date: "2025-08-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(writexl))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(tidyverse))

output = "./output/with_men"
# output = "./output/without_men"

getwd()
knitr::opts_knit$set(root.dir = "C:/Users/brean/Downloads/masters/Atrium/Xerostomia")
getwd()
```

## exploratory analyses:

-   avg taxa detected per sample
-   total taxa detected per group
-   demographics tests (ex. M vs. F) - I've attached an example of a demographics table. You can use MWU or Chi-square for statistical comparisons among the groups

## Read in the file

### With Men

```{r microbiome, echo = FALSE}


## read file 
microbiome_data <- read_excel("csv_files/Xero-Microbiome RA data patient matched AH 07 23 25.xlsx")
microbiome_data <- relocate(microbiome_data, Group, Condition, Site)
microbiome_data <- microbiome_data[,1:length(microbiome_data)-1] %>% as.data.frame()
microbiome_data$Condition[microbiome_data$Condition == "None"] <- ""
```

## Make Microbiome Data without Men

```{r}
microbiome_data <- read_excel("csv_files/Xero-Microbiome RA data patient matched AH 07 23 25.xlsx")
microbiome_data[["XOB ID"]] <- sapply(str_split(microbiome_data$Taxa, "\\."), `[`, 1)


metadata <- read_excel("csv_files/XOB Clinical Data Microbiome + Metatranscriptomics AH 07 23 2025.xlsx") |>
  select(Sex, "XOB ID") |>
  filter(Sex == "F")
metadata[["XOB ID"]] <- paste(sapply(str_split(metadata[["XOB ID"]], "_."), `[`, 1),sapply(str_split(metadata[["XOB ID"]], "_."), `[`, 2),sep="0")


microbiome_data <- merge(microbiome_data, metadata, by = "XOB ID") |> relocate(Group, Condition, Site, Sex, "XOB ID")

microbiome_data$Sex = NULL
microbiome_data[["XOB ID"]] = NULL

write_xlsx(microbiome_data, "csv_files/Xero-Microbiome RA data patient matched AH 07 23 25 females.xlsx", col_names = TRUE)

microbiome_data <- relocate(microbiome_data, Group, Condition, Site)
microbiome_data <- microbiome_data[,1:length(microbiome_data)-1] %>% as.data.frame()
microbiome_data$Condition[microbiome_data$Condition == "None"] <- ""

```

```{r}
## read file 
microbiome_data <- read_excel("csv_files/Xero-Microbiome RA data patient matched AH 07 23 25 females.xlsx")
microbiome_data <- relocate(microbiome_data, Group, Condition, Site)
microbiome_data <- microbiome_data[,1:length(microbiome_data)-1] %>% as.data.frame()
microbiome_data$Condition[microbiome_data$Condition == "None"] <- ""

microbiome_data
```

## Average Amount of Taxa Per Sample

### Average of the Taxa For each Group

### Compute the mean

```{r, echo = FALSE}
avg_taxa_per_sample <- select(microbiome_data, -c(colnames(microbiome_data[,1:5]))) %>% base::apply(MARGIN = 1, FUN = mean) 

avg_taxa_per_sample_df <- cbind(select(microbiome_data, c(colnames(microbiome_data[,1:4]))), data.frame(sample_mean = avg_taxa_per_sample) )
```

### Plot the results

```{r, echo = FALSE}
png("./output/exploratory_analyses/avg taxa per sample- Group.png")

avg_taxa_per_sample_df %>%
  select(Taxa, Group, sample_mean) %>%
  arrange(Group, desc(Group)) %>%
  dplyr::mutate(Taxa = fct_inorder(Taxa) ) %>%
  ggplot(aes(x = Taxa, y = sample_mean)) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  geom_bar(aes(fill = Group), stat = "identity") + 
  labs(x = "")

dev.off()



avg_taxa_per_sample_df$Group_Condition <- paste(avg_taxa_per_sample_df$Group, avg_taxa_per_sample_df$Condition)


png("./output/exploratory_analyses/avg taxa per sample- Group_Condition.png")

avg_taxa_per_sample_df %>%
  select(Taxa, Group_Condition, sample_mean) %>%
  arrange(Group_Condition, desc(Group_Condition)) %>%
  dplyr::mutate(Taxa = fct_inorder(Taxa) ) %>%
  ggplot( aes(y = sample_mean, x = Taxa, fill = Group_Condition)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    labs(fill = "Group Condition", x = "")
dev.off()

rm(avg_taxa_per_sample_df, avg_taxa_per_sample)

```

### Average amount of taxa greater than 0

```{r}
compute_avg <- function(metadata_column){
  
  metadata_column_Var <- sym(metadata_column)
  
    avg_taxa_per_group <- microbiome_data |> select(!!metadata_column_Var, c(colnames(microbiome_data[,6:length(microbiome_data)])))  %>% 
      pivot_longer(-c(!!metadata_column_Var), names_to = "bacteria", values_to = "count" ) |> group_by(!!metadata_column_Var) |> filter(count > 0) |> count(bacteria) 
  
  avg_taxa_per_group$n[avg_taxa_per_group$n != 0] <- 1
  # print(avg_taxa_per_group)
  
  formula <- as.formula(paste("n ~", metadata_column) )    

  avg_taxa_per_group = aggregate(avg_taxa_per_group, formula, FUN = sum)
  
  # avg_taxa_per_group <- microbiome_data |> select(!!metadata_column_Var, c(colnames(microbiome_data[,6:length(microbiome_data)])))  %>% 
  #     pivot_longer(-c(!!metadata_column_Var), names_to = "bacteria", values_to = "avg" )
  # 
  # formula <- as.formula(paste("avg ~", metadata_column) )    
  avg_taxa_per_group$n <- (avg_taxa_per_group$n)/(dim(microbiome_data)[1])
  aggregate(avg_taxa_per_group, formula, FUN = mean) %>% print()
  plot <- aggregate(avg_taxa_per_group, formula, FUN = mean) |>
    
    ggplot(aes(y = n, x=!!metadata_column_Var, fill=!!metadata_column_Var)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    labs(fill = metadata_column, x = "", y = "average") 
  
  png(paste("./output/without_men/exploratory_analyses/bacterial avg of ",metadata_column,".png", sep=""))  
  
   print(plot)
  
  dev.off()
  
}

microbiome_data$Condition[microbiome_data$Condition == ""] <- "Control"
compute_avg("Group")
compute_avg("Site")
compute_avg("Condition")
```

## Total Taxa Detected per Group

### Compute the sum of Group

```{r, echo = FALSE}

compute_sum <- function(metadata_column){
  
  metadata_column_Var <- sym(metadata_column)
  
  amount_taxa_per_group <- microbiome_data |> select(!!metadata_column_Var, c(colnames(microbiome_data[,6:length(microbiome_data)])))  %>% 
      pivot_longer(-c(!!metadata_column_Var), names_to = "bacteria", values_to = "count" ) |> group_by(!!metadata_column_Var) |> filter(count > 0) |> count(bacteria) 
  
  amount_taxa_per_group$n[amount_taxa_per_group$n != 0] <- 1
  print(amount_taxa_per_group)
  
  formula <- as.formula(paste("n ~", metadata_column) )    

  aggregate(amount_taxa_per_group, formula, FUN = sum) |> print()

  plot <- aggregate(amount_taxa_per_group, formula, FUN = sum) |>
    
    ggplot(aes(y = n, x=!!metadata_column_Var, fill=!!metadata_column_Var)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    labs(fill = metadata_column, x = "") 
  
  png(paste(output, "/exploratory_analyses/bacterial sum of ",metadata_column,".png", sep=""))  
  
   print(plot)
  
  dev.off()
  
  }


microbiome_data$Condition[microbiome_data$Condition == ""] <- "Control"
compute_sum("Group")
compute_sum("Site")
compute_sum("Condition")
```

## Demographics tests (ex. M vs. F) -

## I've attached an example of a demographics table. You can use MWU or Chi-square for statistical comparisons among the groups

```{r}
getwd()
demographics <- read_excel("./csv_files/female demographics.xlsx")
demographics
```

## Fishers exact test (small sample size) for gender

```{r}
demographics |> select (Groupa, M, F)
demographics |> select (M, F) |> fisher.test()
```

## Fisher's exact test for Ethnicity

```{r}
demographics[,c(1, grep("Ethnicity", colnames(demographics)))]

demographics[,c(grep("Ethnicity", colnames(demographics)))] |> fisher.test()
```

## Fisher's exact test for COD

```{r}

COD <- demographics[,c(1, grep("COD", colnames(demographics)))]

keys <- lapply(COD[[ colnames(COD)[2] ]], function(s) {
  as.integer(sub(":.*", "", strsplit(s, "/")[[1]]))
})

values <- lapply(COD[[ colnames(COD)[2] ]], function(s) {
  as.integer(sub(".*:", "", strsplit(s, "/")[[1]]))
})

df <- data.frame()
for ( i in range(1:length(keys))) {
  temp <- as.data.frame(t( data.frame(values[[i]]) )) 
  colnames(temp) <- keys[[i]]
  df <- if (ncol(df) == 0) {temp} else {merge(df, temp,  by = intersect(names(df), names(temp)), all = TRUE, sort = F)}
 }

df[is.na(df)] <- 0

print(df)
df |> fisher.test()
```

## T-test for Age

```{r}

age <- read_excel("./csv_files/female demographics.xlsx", sheet = "age")

age <- select(age, control, xerostomia)[-c(1),] 

age
t.test( as.numeric( age$control[!is.na(age$control)] ), as.numeric( age$xerostomia))
```
